quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-10-07 paper-quollr.Rmd
Updated some wordings RMSE to RWBSS

### Compute residuals and Root Mean Square Error (RMSE)

Residuals and root mean squared error (RMSE) are used as goodness of fit metrics for the model. These metrics can be computed using the  `glance()` function, which provides a tidy output for evaluation.

@@

From these per-dimension errors, the function computes absolute error which is the sum of absolute differences across all dimensions and observations and the RMSE which is the average of the total squared error per point.

These metrics are returned in a tibble as `Error` (absolute error) and `RMSE` (root mean squared error). The computation of total absolute error and RMSE is performed in C++ for efficiency using the internal `compute_errors()` function.

@@

Limb muscle cells of mice in @tabula2018 are examined. There are $1067$ single cells, with $14997$ gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \@ref(fig:limb-rmse) (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better \twoD layout.

@@

```{r limb-rmse, fig.cap="Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using RMSE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is perform well at large binwidth (where the binwidth is not enough to capture the data struture) and poorly as bin width decreases. Layout f is the best choice.\\label{fig:limb-rmse}", fig.pos='H', echo=TRUE}

@@

```{r model-limb, fig.cap="Compare the published $2-\\text{D}$ layout (Figure \\ref{fig:limb-rmse}b) and the $2-\\text{D}$ layout selected (Figure \\ref{fig:limb-rmse}f) by RMSE plot (Figure \\ref{fig:limb-rmse}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\\text{-}D$.", fig.pos='H', fig.width=15, fig.height=10, eval=knitr::is_latex_output()}

@@

```{r model-limb-html, eval=knitr::is_html_output(), fig.cap="Compare the published $2-\\text{D}$ layout (Figure \\@ref(fig:limb-rmse) b) and the $2-\\text{D}$ layout selected (Figure \\@ref(fig:limb-rmse) f) by RMSE plot (Figure \\@ref(fig:limb-rmse)) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\\text{-}D$."}

@@

In contrast to the common practice of visually inspecting scatterplots for clusters or patterns, `quollr` provides a quantitative route for evaluation. It enables the computation of RMSE and residuals between the original high-dimensional data and the lifted model, offering interpretable diagnostics. These diagnostics are complemented by interactive linked plots and high-dimensional dynamic visualizations using the `langevitour` package, allowing users to inspect where the model fits well and where it does not.

To support efficient computation, particularly for large-scale datasets, several core functions in `quollr` are implemented in C++ using `Rcpp` and `RcppArmadillo`. These include functions for computing Euclidean distances in high-dimensional and \twoD space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, `compute_highd_dist()` accelerates nearest neighbor lookup in high-dimensional space, `compute_errors()` calculates RMSE and total absolute error efficiently, and `calc_2d_dist_cpp()` speeds up distance calculations in \twoD. Additionally, `gen_hex_coord_cpp()` constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.

>>>>

### Compute residuals and Root Within Bin Sum of Square (RWBSS)

Root Within Bin Sum of Square (RWBSS) are used as goodness of fit metrics for the model. These metrics can be computed using the  `glance()` function, which provides a tidy output for evaluation.

@@

From these per-dimension errors, the function computes absolute error which is the sum of absolute differences across all dimensions and observations and the RWBSS which is the average of the total squared error per point.

These metrics are returned in a tibble as `Error` (absolute error) and `RWBSS` (root mean squared error). The computation of total absolute error and RWBSS is performed in C++ for efficiency using the internal `compute_errors()` function.

@@

Limb muscle cells of mice in @tabula2018 are examined. There are $1067$ single cells, with $14997$ gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \@ref(fig:limb-rwbss) (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better \twoD layout.

@@

```{r limb-rwbss, fig.cap="Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using RWBSS for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is perform well at large binwidth (where the binwidth is not enough to capture the data struture) and poorly as bin width decreases. Layout f is the best choice.\\label{fig:limb-rwbss}", fig.pos='H', echo=TRUE}

@@

```{r model-limb, fig.cap="Compare the published $2-\\text{D}$ layout (Figure \\ref{fig:limb-rwbss}b) and the $2-\\text{D}$ layout selected (Figure \\ref{fig:limb-rwbss}f) by RWBSS plot (Figure \\ref{fig:limb-rwbss}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\\text{-}D$.", fig.pos='H', fig.width=15, fig.height=10, eval=knitr::is_latex_output()}

@@

```{r model-limb-html, eval=knitr::is_html_output(), fig.cap="Compare the published $2-\\text{D}$ layout (Figure \\@ref(fig:limb-rwbss) b) and the $2-\\text{D}$ layout selected (Figure \\@ref(fig:limb-rwbss) f) by RWBSS plot (Figure \\@ref(fig:limb-rwbss)) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\\text{-}D$."}

@@

In contrast to the common practice of visually inspecting scatterplots for clusters or patterns, `quollr` provides a quantitative route for evaluation. It enables the computation of RWBSS and residuals between the original high-dimensional data and the lifted model, offering interpretable diagnostics. These diagnostics are complemented by interactive linked plots and high-dimensional dynamic visualizations using the `langevitour` package, allowing users to inspect where the model fits well and where it does not.

To support efficient computation, particularly for large-scale datasets, several core functions in `quollr` are implemented in C++ using `Rcpp` and `RcppArmadillo`. These include functions for computing Euclidean distances in high-dimensional and \twoD space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, `compute_highd_dist()` accelerates nearest neighbor lookup in high-dimensional space, `compute_errors()` calculates RWBSS and total absolute error efficiently, and `calc_2d_dist_cpp()` speeds up distance calculations in \twoD. Additionally, `gen_hex_coord_cpp()` constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.

### 2025-10-07 paper-quollr.tex
Updated some wordings RMSE to RWBSS

\subsubsection{Compute residuals and Root Mean Square Error (RMSE)}\label{compute-residuals-and-root-mean-square-error-rmse}

Residuals and root mean squared error (RMSE) are used as goodness of fit metrics for the model. These metrics can be computed using the \texttt{glance()} function, which provides a tidy output for evaluation.

@@

From these per-dimension errors, the function computes absolute error which is the sum of absolute differences across all dimensions and observations and the RMSE which is the average of the total squared error per point.

These metrics are returned in a tibble as \texttt{Error} (absolute error) and \texttt{RMSE} (root mean squared error). The computation of total absolute error and RMSE is performed in C++ for efficiency using the internal \texttt{compute\_errors()} function.

@@

Limb muscle cells of mice in \citet{tabula2018} are examined. There are \(1067\) single cells, with \(14997\) gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \ref{fig:limb-rmse} (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better $2\text{-}D$ layout.

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/limb-rmse-1} \caption{Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using RMSE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is perform well at large binwidth (where the binwidth is not enough to capture the data struture) and poorly as bin width decreases. Layout f is the best choice.\label{fig:limb-rmse}}\label{fig:limb-rmse}
\end{figure}

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/model-limb-1} \caption{Compare the published $2-\text{D}$ layout (Figure \ref{fig:limb-rmse}b) and the $2-\text{D}$ layout selected (Figure \ref{fig:limb-rmse}f) by RMSE plot (Figure \ref{fig:limb-rmse}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\text{-}D$.}\label{fig:model-limb}
\end{figure}

@@

In contrast to the common practice of visually inspecting scatterplots for clusters or patterns, \texttt{quollr} provides a quantitative route for evaluation. It enables the computation of RMSE and residuals between the original high-dimensional data and the lifted model, offering interpretable diagnostics. These diagnostics are complemented by interactive linked plots and high-dimensional dynamic visualizations using the \texttt{langevitour} package, allowing users to inspect where the model fits well and where it does not.

To support efficient computation, particularly for large-scale datasets, several core functions in \texttt{quollr} are implemented in C++ using \texttt{Rcpp} and \texttt{RcppArmadillo}. These include functions for computing Euclidean distances in high-dimensional and $2\text{-}D$ space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, \texttt{compute\_highd\_dist()} accelerates nearest neighbor lookup in high-dimensional space, \texttt{compute\_errors()} calculates RMSE and total absolute error efficiently, and \texttt{calc\_2d\_dist\_cpp()} speeds up distance calculations in $2\text{-}D$. Additionally, \texttt{gen\_hex\_coord\_cpp()} constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.

>>>>

\subsubsection{Compute residuals and Root Within Bin Sum of Square (RWBSS)}\label{compute-residuals-and-root-within-bin-sum-of-square-rwbss}

Root Within Bin Sum of Square (RWBSS) are used as goodness of fit metrics for the model. These metrics can be computed using the \texttt{glance()} function, which provides a tidy output for evaluation.

@@

From these per-dimension errors, the function computes absolute error which is the sum of absolute differences across all dimensions and observations and the RWBSS which is the average of the total squared error per point.

These metrics are returned in a tibble as \texttt{Error} (absolute error) and \texttt{RWBSS} (root mean squared error). The computation of total absolute error and RWBSS is performed in C++ for efficiency using the internal \texttt{compute\_errors()} function.

@@

Limb muscle cells of mice in \citet{tabula2018} are examined. There are \(1067\) single cells, with \(14997\) gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \ref{fig:limb-rwbss} (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better $2\text{-}D$ layout.

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/limb-rwbss-1} \caption{Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using RWBSS for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is perform well at large binwidth (where the binwidth is not enough to capture the data struture) and poorly as bin width decreases. Layout f is the best choice.\label{fig:limb-rwbss}}\label{fig:limb-rwbss}
\end{figure}

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/model-limb-1} \caption{Compare the published $2-\text{D}$ layout (Figure \ref{fig:limb-rwbss}b) and the $2-\text{D}$ layout selected (Figure \ref{fig:limb-rwbss}f) by RWBSS plot (Figure \ref{fig:limb-rwbss}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\text{-}D$.}\label{fig:model-limb}
\end{figure}

@@

In contrast to the common practice of visually inspecting scatterplots for clusters or patterns, \texttt{quollr} provides a quantitative route for evaluation. It enables the computation of RWBSS and residuals between the original high-dimensional data and the lifted model, offering interpretable diagnostics. These diagnostics are complemented by interactive linked plots and high-dimensional dynamic visualizations using the \texttt{langevitour} package, allowing users to inspect where the model fits well and where it does not.

To support efficient computation, particularly for large-scale datasets, several core functions in \texttt{quollr} are implemented in C++ using \texttt{Rcpp} and \texttt{RcppArmadillo}. These include functions for computing Euclidean distances in high-dimensional and $2\text{-}D$ space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, \texttt{compute\_highd\_dist()} accelerates nearest neighbor lookup in high-dimensional space, \texttt{compute\_errors()} calculates RWBSS and total absolute error efficiently, and \texttt{calc\_2d\_dist\_cpp()} speeds up distance calculations in $2\text{-}D$. Additionally, \texttt{gen\_hex\_coord\_cpp()} constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.
